---
title: "Peer-graded Assignment: Statistical Inference Course Project"
author: "Irina Z"
date: "9/22/2020"
output: pdf_document
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

This report contains two parts of the course project:  
1. A simulation exercise  
2. Basic inferential data analysis  

## Part 1: Simulation Exercise

### Synopsis

In this report, I investigate the **exponential distribution** in R and compare it with the **Central Limit Theorem**. 
\  
The exponential distribution will be simulated in R with `rexp(n, lambda)` where lambda is the rate parameter. The mean of exponential distribution is $\frac{1}{\lambda}$, and the standard deviation is also $\frac{1}{\lambda}$. We set $\lambda = 0.2$ for all of the simulations and investigate the distribution of averages of 40 exponentials and do a thousand simulations.

##### Question 1 : Show the sample mean and compare it to the theoretical mean distribution  
First, I run simulation:
```{r sample_gen, echo=TRUE, include=TRUE}
#  Setting the seed for reproducability
set.seed(1956)
n <- 40 
n_sims <- 1000
lambda <- 0.2

## Simulate the sample
SampleMean <- NULL
for(i in 1:n_sims) {
  SampleMean <- c(SampleMean, mean(rexp(n, lambda)))
}
```
  
The theoretical mean distribution is $E[X]=\frac{1}{\lambda}=\frac{1}{0.2}=5$, and the sample mean is $\overline{X}=$ `r mean(SampleMean)`, which is **close to the theoretical mean distribution**.  
Difference between the values is:
```{r mean_diff, echo=TRUE, include=TRUE}
abs((1/lambda) - mean(SampleMean))
```
  
  
We can also show how close the sample mean and the theoretical mean are on the plot which is shown in the Appendix (Picture 1).


##### Question 2 : Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution
The sample variance is:
```{r sample_var, echo=TRUE, include=TRUE}
SampleVar <- var(SampleMean)
```

Which is equal to `r SampleVar`.
  
The variance of the sample mean is $var(\overline{X})=\frac{\sigma^2}{n}$, where ${\sigma}$ is the standard deviation equal to  $\sigma=\frac{1}{\lambda}=\frac{1}{0.2}=5$. Therefore, the theoretical variance is $var(\overline{X})=\frac{5^2}{40}=\frac{25}{40}=0.625$. 
  
  
Difference between the values is:
```{r var_diff, echo=TRUE, include=TRUE}
abs((1/lambda)^2/n - var(SampleMean))
```

This difference is also small, so **the variance of the sample mean and the theoretical variance are close**.


##### Question 3 : Show that the distribution is approximately normal
To show that the distribution is approximately normal, I plot histogram with density function:
```{r plot_norm, echo=TRUE, include=TRUE}
ggplot(data.frame(SampleMean), aes(x = SampleMean)) +
        geom_histogram(bins = 40, color = "darkseagreen", fill = "honeydew2",
                       aes(y = ..density..))+
            stat_function(fun = dnorm, args = list(mean = mean(SampleMean),
                                                   sd = sd(SampleMean)), 
                          aes(color = "Normal Distribution"), size = 1.5) +
            stat_density(geom = "line", aes(color = "Sample Distribution"), size = 1.5)  +
            labs(x = "Exponential Mean", y = "Density",
                 title = "Normal Distribution Comparision") +
        scale_color_manual(name = "Distributions", 
                             values = c("Normal Distribution" = "red", 
                                        "Sample Distribution" = "blue")) +
        theme_bw()+
        theme(legend.position = c(0.85, 0.85),
              legend.background = element_rect(colour = 'grey'),
              plot.title = element_text(hjust = 0.5)) 
```

This plot indicates that density curve of our sample ($\color{blue}{blue\;line}$) is similar to normal distribution curve ($\color{red}{red\;line}$). Therefore, **the distribution is approximately normal**.
\  

We can also use a **quantile-quantile (Q-Q) plot** of sample mean against a normal distribution. If the data points follow the linearity represented by the normal distribution, then we can say that the data have normal distribution.
\  

The Q-Q Normal Plot is shown in the Appendix (Picture 2). This plot also **indicates the normal distribution of the exponential data**.

\newpage

## Appendix to Part 1

#### Picture 1. Theoretical and sample means:
```{r mean_gplot, echo=TRUE, include=TRUE}
ggplot(mapping = aes(SampleMean)) +
        geom_histogram(bins = 40, color = "darkseagreen", fill = "honeydew2") +
        labs(x = "Exponential Mean", y = "Frequency", 
             title = "Exponential Mean Distribution") +
        geom_vline(aes(xintercept = 5, color = "Theoretical mean"), size = 1) +
        geom_vline(aes(xintercept = mean(SampleMean), color = "Sample mean"), size = 1) +
        scale_color_manual(name = "Means", 
                             values = c("Theoretical mean" = "red", 
                                        "Sample mean" = "blue")) +
        theme_bw()+
        theme(legend.position = c(0.85, 0.85),
              legend.background = element_rect(colour = 'grey'),
              plot.title = element_text(hjust = 0.5)) 
```

\  
Theoretical mean is shown by the $\color{red}{red\;line}$, while the sample mean is shown by the $\color{blue}{blue\;line}$. We can see that the two lines are close, therefore, **the values of the sample mean and the theoretical mean are close**.

\newpage

#### Picture 2. Comparison to normal distribution via Q-Q plot:

```{r qq_gplot, echo=TRUE, include=TRUE}
ggplot(data.frame(SampleMean), aes(sample = SampleMean, col = "darkseagreen")) + 
        geom_qq() + 
        stat_qq_line() +
        labs(x = "Normal Theoretical Quantiles", y = "Exponential Data Quantiles",
                 title = "Normal Distribution Comparision via Q-Q Plot") +
        theme_bw() +
        theme(legend.position = "none",
              plot.title = element_text(hjust = 0.5)) 
```

\  
The data points on the Q-Q plot have outliers in the bottom left and upper right corners, but otherwise the linearity of the points suggests that **the data are approximately normally distributed**.